---
title: "MA678 Final Report - Suicide"
author: "Yuchen Huang"
date: "2023-12-12"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lme4)
library(rstanarm)
library(coefplot)
library(kableExtra)
library(performance)
set.seed(678)

suicide <- read.csv("master.csv")
hdi <- read.csv("HDR21-22_Composite_indices_complete_time_series.csv")
hdi$country[hdi$country == "Korea (Republic of)"] <- "Republic of Korea"
```

## ABSTRACT

## INTRODUCTION

Suicide is a major public mental health issue globally, affecting
individuals across various socio-economic backgrounds. It is a complex
phenomenon that could potentially be linked to a multitude of factors
including economic conditions, societal norms, and individual
circumstances.

Here I want to focus on finding out the relationship between suicide
rate and HDI(Human Development Index). A high HDI represents a high
development, and it is calculated by four variables:

-   `LE`: Life Expectancy

-   `EI`: Education Index, and this is the mean of mean year schooling
    and expected year schooling

-   `GNI per capita`: Gross National Income per capita

$$
HDI = (\frac{LE-20}{85-20}\times\frac{MYSI+EYSI}{2}\times\frac{ln(GNIpc)-ln(100)}{ln(20,000)-ln(100)})^{\frac{1}{3}}
$$

Also, I will consider the gender and generations in the modeling, and I
will explain the data more detailed in the following sections.

## METHOD

#### 1. Data Sources

The data I used is a combination from two sources, I got the suicide
rate from 1985 to 2016 from Kaggle and HDI with related index from
UNDP(United Nations Development Programme) from 1990 to 2022. As the
suicide data has many missing values in 2016, I selected data from
1990-2015 to analyse.

#### 2. Data cleaning

The datasets are from different websites, the countries in the two
datasets are different. So The first step of data cleaning is to find
out the same countries shared in them.

```{r, echo=F}
##get only hdi from the data
HDI <- hdi |>
  select(country, starts_with("hdi_")) |>
  select(-hdi_rank_2021)

## remove NA
HDI <- HDI |>
  filter(is.na(hdi_1990)==F)

## remove hdi in the column names
colnames(HDI) <- gsub("hdi_", "", colnames(HDI)) 
HDI <- HDI |> select(-starts_with("f"), -starts_with("m"))

## convert HDI into long format
HDI_long <- HDI |>
  pivot_longer(cols = -country, names_to = "year", values_to = "HDI")

## select suicide data after 1990
suicide <- suicide |>
  filter(year >= 1990)
HDI_long$year <- as.integer(HDI_long$year)

## combine HDI_long and suicide together
Suicide <- suicide |>
  left_join(HDI_long, by = c("country", "year")) |>
  mutate(HDI = HDI)
Suicide <- Suicide |>
  select(-HDI.for.year) |>
  relocate(HDI, .before = gdp_for_year....)

## check how many lines HDI is NA
Suicide_clean <- Suicide |>
  filter(is.na(HDI)==F)
#unique(Suicide_na$country)
paste0("There are ",length(unique(Suicide_clean$country))," countries in the cleaning data set, and all the large countries are inclued, like ", unique(Suicide_clean$country)[5]," and ", unique(Suicide_clean$country)[10], ".")
```

Then I add the HDI, LE, MSY, ESY, and GNI per capita to the cleaning
data set.

```{r, echo = F}
##Add Life Expectancy, MYSI, EYSI, GNI per capita
LE <- hdi |>
  select(country, starts_with("le_19"), starts_with("le_20"))

colnames(LE) <- gsub("le_", "", colnames(LE))

LE <- LE |>
  filter(is.na(`1990`)==F)

LE_long <- LE |>
  pivot_longer(cols = -country, names_to = "year", values_to = "LE")

LE_long$year <- as.integer(LE_long$year)

Suicide_clean <- Suicide_clean |>
  left_join(LE_long, by = c("country", "year")) |>
  mutate(LE = LE)|>
  relocate(LE, .before = gdp_for_year....)

EYS <- hdi |>
  select(country, starts_with("eys_19"), starts_with("eys_20"))
colnames(EYS) <- gsub("eys_", "", colnames(EYS))
EYS <- EYS |>
  filter(is.na(`1990`)==F)

EYS_long <- EYS |>
  pivot_longer(cols = -country, names_to = "year", values_to = "EYS")
EYS_long$year <- as.integer(EYS_long$year)

Suicide_clean <- Suicide_clean |>
  left_join(EYS_long, by = c("country", "year")) |>
  mutate(EYS = EYS) |>
  relocate(EYS, .before = gdp_for_year....)

MYS <- hdi |>
  select(country, starts_with("mys_19"), starts_with("mys_20"))
colnames(MYS) <- gsub("mys_", "", colnames(MYS))
MYS <- MYS |>
  filter(is.na(`1990`)==F)

MYS_long <- MYS |>
  pivot_longer(cols = -country, names_to = "year", values_to = "MYS")
MYS_long$year <- as.integer(MYS_long$year)

Suicide_clean <- Suicide_clean |>
  left_join(MYS_long, by = c("country", "year")) |>
  mutate(MYS = MYS) |>
  relocate(MYS, .before = gdp_for_year....)

GNI <- hdi |>
  select(country, starts_with("gnipc_19"), starts_with("gnipc_20"))
colnames(GNI) <- gsub("gnipc_", "", colnames(GNI))
GNI <- GNI |>
  filter(is.na(`1990`)==F)

GNI_long <- GNI |>
  pivot_longer(cols = -country, names_to = "year", values_to = "GNI")
GNI_long$year <- as.integer(GNI_long$year)

Suicide_clean <- Suicide_clean |>
  left_join(GNI_long, by = c("country", "year")) |>
  mutate(GNI = GNI) |>
  relocate(GNI, .before = gdp_for_year....)

```

```{r, echo = FALSE, warning = FALSE}
Suicide_clean <- Suicide_clean |>
  select(-country.year, -gdp_for_year...., suicides_no, -age)
## cleaning environment
rm(hdi, HDI, HDI_long) 
rm(suicide, Suicide, Suicide_na)
rm(LE, LE_long)
rm(MYS, MYS_long, EYS, EYS_long)
rm(GNI, GNI_long)
```

However, the data from UNDP is the actual value instead of index, and
they are in different scales, so I make all the imput into a scales like
HDI(around 1). For LE, MYS, EYS, and GNI per capita, I directly use the
formula above to transform them into a decimal less than 1.

```{r, echo = F}
Suicide_clean$LE <- (Suicide_clean$LE-20)/(85-20)
Suicide_clean$EI <- (Suicide_clean$MYS+Suicide_clean$EYS)/2
Suicide_clean$GNI <- (log(Suicide_clean$GNI)-log(100))/(log(20000)-log(100))
Suicide_clean$population <-  Suicide_clean$population/100000
Suicide_clean$gdp_per_capita.... <- log(Suicide_clean$gdp_per_capita....)/10

Suicide_clean <- Suicide_clean |>
  relocate(EI, .after = LE) |>
  select(-MYS, -EYS)
```

#### 3. Data Introduction

Figure1. is the first six rows of the cleaning data, and this is a long
format data. Each row represents a group of people with a same gender
from different countries in different years. The `suicides.100k.pop`
means the ratio of $\frac{SuicideNum}{Population}\times10^6$ of the
group.

Figure 1.

```{r, echo = F}
kable(head(Suicide_clean[,c(1:3,6:10,12)]))
```

In figure2. here's the generation types I mention in the introduction
part, this category is very important in social science analysis, which
is why I choose to use it rather the age groups.

Figure2.

```{r, echo = F}
kable(table(Suicide_clean$generation))
```

#### 4. EDA

Figure 3. and 4. are a bar plot and a scatter considering different
genders. It is obvious that female has a lower suicide rate than the
male. This indicates that gender can be a significant variable in the
models.

Figure 3.&4.

```{r, echo = F}
Suicide_clean |>
  ggplot(mapping = aes(HDI,(suicides.100k.pop))) +
  geom_boxplot(aes(fill = sex), outlier.color = "red", outlier.shape = 16)

Suicide_clean |>
  ggplot() +
  aes(x = HDI, y = suicides.100k.pop, color = sex) +
  geom_point(size = .5)+
  geom_smooth(method = "lm")
```

Figure5. is showing random 12 countries' data considering time versus
suicide rate, with different colors representing different generations.

It is be observed that different countries have very different trend of
suicide rates, and people from different genders might vary very much.

As a result, the generation and country will be two significant inputs
as well.

Figure 5.

```{r, echo = F}
Suicide_clean |> 
  subset(country %in% sample(Suicide_clean$country, 12)) |>
  ggplot() +
  aes(x = year, y = suicides.100k.pop, color = generation) +
  geom_point(size = .5) +
  facet_wrap(~country)
```

#### 5. Models

a)  Null Model

-   Model

$$
Suiceds.100k.pop = \beta_0
$$

```{r, echo = F}
null_model <- mean(Suicide_clean$suicides.100k.pop)
```

-   Residual Plot

Figure 6.

```{r, echo = F}
null_pred <- Suicide_clean |> select(country, suicides.100k.pop)
null_pred$predict <- rep(null_model, times = nrow(Suicide_clean))
null_pred$resid <- null_pred$suicides.100k.pop -null_pred$predict

ggplot(null_pred, aes(x = predict, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Plot of Null Mofrl", x = "Predicted Suicide Rate", y = "Residuals") +
  theme_minimal()
```

b)  Complete Pooling

-   Model

$$
Suiceds.100k.pop = \beta_{0}+\beta_1x_{year} + \beta_2x{sex} + \beta_3x_{population} + \beta_4{HDI} + \beta_5x_{LE} + 
$$

$$
\beta_{6}x_{EI} + \beta_{7}x_{GNI} + \beta_{8}x_{gdp_per_capita} + \beta_{9}x_{generation} + \beta_{10}x_{country}
$$

```{r, echo = F}
comp_model <- Suicide_clean |>
  lm(formula = suicides.100k.pop ~ year + sex + population + HDI + LE + EI + GNI + gdp_per_capita.... + generation + country)
#summary(comp_model)
```

-   Residual Plot

Figure 7.

```{r, echo = F}
plot(comp_model,which =1)
```

c)  No Pooling

-   Model

$$
Suiceds.100k.pop = \beta_{i}+\beta_{i1}x_{year} + \beta_{i2}x{sex} + \beta_{i3}x_{population} + \beta_{i4}{HDI} + \beta_{i5}x_{LE} + 
$$

$$
\beta_{6i}x_{EI} + \beta_{7i}x_{GNI} + \beta_{8i}x_{gdp_per_capita} + \beta_{9i}x_{generation} + \beta_{10i}x_{country}
$$

```{r, echo = F}
no_model <- Suicide_clean %>%
  split(.$country) %>%
  lapply(function(df) {
    lm(suicides.100k.pop ~ year + sex + population + HDI + LE + EI + GNI + gdp_per_capita.... + generation, data = df)
  })
```

-   Visualization

Figure 8.

```{r, echo = F}
Suicide_clean |> 
  subset(country %in% sample(Suicide_clean$country, 12)) |>
  ggplot() +
  aes(x = year, y = suicides.100k.pop, color = generation) +
  geom_point(size = .5) +
  facet_wrap(~country) +
  geom_smooth(method = "lm", color = "blue", linewidth = .5)
```

d)  Partial Pooling

-   Model $$
    Suiceds.100k.pop = \beta_{0i}+\beta_1x_{year} + \beta_2x{sex} + \beta_3x_{population} + \beta_4{HDI} + \beta_5x_{LE} + 
    $$

$$
\beta_{6}x_{EI} + \beta_{7}x_{GNI} + \beta_{8}x_{gdp_per_capita} + \beta_{9}x_{generation} + \beta_{10}x_{country}
$$

```{r, echo = F}
partial_model <- Suicide_clean |>
  lmer(formula = suicides.100k.pop ~ year + sex + population + HDI + LE + EI + GNI + gdp_per_capita.... + generation + (1|country))
#summary(partial_model)
```

-   Coefficient Plot

Figure 9.

```{r, echo = F}
coefplot(partial_model)
```

```{r, echo = F, warning=F}
selected_countries <- sample(unique(Suicide_clean$country), 12)

selected_data <- Suicide_clean |>
  filter(country %in% selected_countries)

partial_model_selected <- 
  lmer(suicides.100k.pop ~ year + sex + population + HDI + LE + EI + GNI + gdp_per_capita.... + generation + (1|country), data = selected_data)


fixef_estimates_selected <- fixef(partial_model_selected)
ranef_estimates_selected <- ranef(partial_model_selected)$country
ranef_estimates_selected$`(Intercept)` <- ifelse(ranef_estimates_selected$`(Intercept)` <0,0,ranef_estimates_selected$`(Intercept)`)

country_intercepts_selected <- 
  data.frame(country = rownames(ranef_estimates_selected), intercept = ranef_estimates_selected[, "(Intercept)"])

plot_data_selected <- selected_data |>
  left_join(country_intercepts_selected, by = "country")


ggplot(plot_data_selected, aes(x = year, y = suicides.100k.pop, group = country)) +
  geom_point(aes(color = generation), size = 0.5) +
  geom_line(aes(y = intercept + fixef_estimates_selected["year"]), color = "blue", size = 0.5) +
  facet_wrap(~country) +
  labs(title = "Varying Intercept with Shared Slope for Selected Countries") +
  theme_minimal()
```

-   Residual Plot

```{r, echo = F}
partial_pred <- Suicide_clean |> select(country, suicides.100k.pop)
partial_pred$predicted_suicide_rate <- predict(partial_model, newdata = Suicide_clean)
partial_pred$predicted_suicide_rate <- ifelse(partial_pred$predicted_suicide_rate<=0, 0, partial_pred$predicted_suicide_rate)

partial_pred$residuals <- partial_pred$suicides.100k.pop - partial_pred$predicted_suicide_rate

partial_pred |>
  ggplot(aes(x = predicted_suicide_rate, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Plot of partial pooling", x = "Predicted Suicide Rate", y = "Residuals") +
  theme_minimal()
```

## RESULT

Here I will analyze how well these four models fit by calculate the
$R^2$ , as the $R^2$ approaches to 1, it means that the model fits the
model more; while $R^2$ approaches to 0, it means the model doesn't have
much correlation with the data

```{r, echo = F}
r2_comp <- r2(comp_model)$R2
r2_partial <- r2(partial_model)$R2

r2_list <- lapply(no_model, function(model) {
  summary(model)$r.squared
})
r2_values <- unlist(r2_list)
r2_no <- sum(r2_values)/74

```

By calculation, I get the values of $R^2$ for the multilevel models.

-   Partial pooling: 0.552

-   Complete pooling 0.530

-   No pooling: 0.651

It seems that the No pooling model fits the best, and the complete
pooling explained the model worst. However, all of them has a $R^2$
larger than 0.5, this can somewhat show that all of these 3 voters can
express the data.

## DISCUSSION

It seems that all the models

Also, through the residual plots shows that there's some trend, so their
might be some missing predictors that I haven't considered. This also
might be caused by some correlation among the variables. However, based
on what we learned so far, I cannot solve the problems yet.

## Reference

[Human Development
Reports](https://hdr.undp.org/data-center/documentation-and-downloads)

[Kaggle](https://www.kaggle.com/datasets/russellyates88/suicide-rates-overview-1985-to-2016/data)

<https://hdr.undp.org/data-center/country-insights#/ranks>

<https://databank.worldbank.org/source/world-development-indicators>

<https://www.who.int/health-topics/suicide#tab=tab_1>
